---
title: "Insee population estimate per department / year"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen=999)
data_storage_url <- "/media/lagrange/homes/lagrange_barthelemy/_DATA_STORE/"
```

```{r loadLibraries, message=FALSE, warning=FALSE, include=FALSE}
library(data.table)
library(dplyr)
library(tidyr)
library(xlsx)
library(kableExtra)
library(ggplot2)
```



# Introduction

The purpose of this report is to prepare a dataset containing France population
estimate per department.
Official statistical burea (https://insee.fr) distribute these data but in an 
ugly format.

See https://www.insee.fr/fr/statistiques/1893198 for details.

> Chaque année, l'Insee estime la population des régions et des départements 
(France métropolitaine et DOM) à la date du 1ᵉʳ janvier. Ces estimations 
annuelles de population sont déclinées par sexe et par âge (quinquennal, 
classes d'âge).


# Data sources

In order to creat this dataset we will combine several sources. First from
insee.fr we will get official communes, departements and region lists with 
their insee codes and names ([insee.fr](https://www.insee.fr/fr/information/3720946) ).

We will combine these data and add the official zip codes from La Poste (French
Mail).

```{r dataSources}
# https://www.insee.fr/fr/statistiques/fichier/1893198/estim-pop-dep-sexe-gca-1975-2020.xls
population_france_url <- paste0(
  data_storage_url
  , "insee.fr/population/raw/estim-pop-dep-sexe-gca-1975-2020.xls"
) 
```


# Get data

The data is badly formatted. First it's an excel fils with multiple tabs, one
per year actually and it's in long format with subtotals as line and columns,
not to mentio merged cells.

## Get yearly data (function)

We make a function to parse a specific sheet and will loop to get the whole data
set.

```{r loadInseeMapping}
get_insee_year <- function(
  year = 1975
  , url = population_france_url){
  message(year)
  my_data <- xlsx::read.xlsx(
    file = url
    , sheetName = as.character(year)
    , startRow =  5
  ) %>% 
    as.data.table()
  ## keep only first 20 columns
  my_data <- my_data[ , 1:20]
  # Rename Features
  setnames(my_data, c(
    "code_departement"
    , "nom_departement"
    , "total_age_0_a_19", "total_age_20_a_39", "total_age_40_a_59", "total_age_60_a_74", "total_age_75_plus", "total_total"
    , "homme_age_0_a_19", "homme_age_20_a_39", "homme_age_40_a_59", "homme_age_60_a_74", "homme_age_75_plus", "homme_total"
    , "femme_age_0_a_19", "femme_age_20_a_39", "femme_age_40_a_59", "femme_age_60_a_74", "femme_age_75_plus", "femme_total"
  )
  )
  # Remove unecessary lines
  if(year <= 1989)  {
    my_data <- my_data[ -c(97:100) ]
  }
  if(year > 1989 & year < 2014) {
    my_data <- my_data[ -c(97) ]
    my_data <- my_data[ 1:100 ]
  }
  if(year >= 2014) {
    my_data <- my_data[ -c(97) ]
    my_data <- my_data[ 1:101 ]
  }
  # Reshape to long format
  my_data <- data.table::melt(my_data, id.vars = c("code_departement", "nom_departement"))
  
  # adjust variables
  my_data[ , ':=' (
    sex = character()
    , categorie = character()
    , annee = year
    , population = value
    , value = NULL
  )
  ]
  # Assign features
  my_data[ grep(pattern = "homme", x = variable), sex := "homme"]
  my_data[ grep(pattern = "femme", x = variable), sex := "femme"]
  my_data[ grep(pattern = "total_age", x = variable), sex := "total"]
  my_data[ grep(pattern = "total_total", x = variable), sex := "total"]
  my_data[ grep(pattern = "age_0_a_19", x = variable), categorie := "age_0_a_19"]
  my_data[ grep(pattern = "age_20_a_39", x = variable), categorie := "age_20_a_39"]
  my_data[ grep(pattern = "age_40_a_59", x = variable), categorie := "age_40_a_59"]
  my_data[ grep(pattern = "age_60_a_74", x = variable), categorie := "age_60_a_74"]
  my_data[ grep(pattern = "age_75_plus", x = variable), categorie := "age_75_plus"]
  my_data[ grep(pattern = "age_75_plus", x = variable), categorie := "age_75_plus"]
  my_data[ grep(pattern = "_total$", x = variable), categorie := "total"]
  my_data[ , variable := NULL]
  my_data
}
```

## Test on some years

Let's test this on 2020 and 1975

```{r}
population_2020 <- get_insee_year(url = population_france_url, year = 2020)
sum(population_2020[ sex == "total" & categorie == "total"]$population)
population_2016 <- get_insee_year(url = population_france_url, year = 2016)
sum(population_2016[ sex == "total" & categorie == "total"]$population)
population_2000 <- get_insee_year(url = population_france_url, year = 2000)
sum(population_2000[ sex == "total" & categorie == "total"]$population)
population_1975 <- get_insee_year(url = population_france_url, year = 1975)
sum(population_1975[ sex == "total" & categorie == "total"]$population)
```

## Get all years

```{r}
france_population_per_department <- data.table::rbindlist(
  lapply(X = 1975:2020, FUN = get_insee_year)
  )
```

# Visualisation   

```{r}
g <- ggplot(data = france_population_per_department[ categorie == "total" & sex == "total"])
g <- g + geom_line(aes(x= annee, y = population))
g <- g + facet_wrap(facets = . ~ code_departement, ncol = 10, scales = "free_y")
g <- g + theme_minimal() +
    theme(
      #legend.position = "none"
      strip.text.x = element_text(
        size=5
        , face = "bold"
        , vjust = -0.0,
        margin = margin(-0.00, 0, 0, 0, "cm")
      )
      , strip.background = element_rect(
        color ="white"
        , fill ="white"
        , size = 0.5
        #, linetype="solid"
      )
      , axis.text.x = element_text(
        angle = 90
        , size = 8
      )
      )
g
```


# Save data

```{r}
save(
  france_population_per_department
  , file = paste0(
    data_storage_url
    , "insee.fr/population/raw/estim-pop-dep-sexe-gca-1975-2020.Rda"
  )
)
```

