---
title: "Mortality in France, major events"
output: html_document
---
    
    ```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r loadLibraries, message=FALSE, warning=FALSE, include=FALSE}
library(geojsonio) # require protonlite install deb : libprotobuf-dev



library(geojsonsf)

library(data.table)
library(dplyr)

library(mapproj)
library(kableExtra)
# Need a recent version of V8
# https://github.com/jeroen/V8
# Ubuntu Xenial (16.04) and Bionic (18.04) only
# sudo add-apt-repository ppa:cran/v8
# sudo apt-get update
# sudo apt-get install libnode-dev
#devtools::install_github("jeroen/V8")
library(rmapshaper)
library(broom)
library(ggplot2)
library(ggthemes)
```



# Introduction

The purpose of this report is to prepare a dataset containing France 
geographic boundaries at 'commune' level (city / county level).

We will import available opendata geojson file and prepare it for plotting
with ggplot and various operations.

# Data sources

There are several version of this dataset available, they integrate progressively
the evolution of French administrative boundaries. However, we use an older
version to get details of Paris & Marseilles districts (arrondissements).

Source of this data is https://github.com/gregoiredavid/france-geojson
Eacch commune is associated with it's insee code.

```{r dataSources}
# France shapefile from data.gouv.fr
# don't use this version as there are no detials for Paris districts
# france_sp_url <- "https://data.datapleth.io/ext/france/spatial/communes-simple/communes-20190101.json"

## Important to use this version to get communes inside Paris & Marseilles.
france_geojson <- "https://github.com/gregoiredavid/france-geojson/raw/v2.0.2/communes.geojson"
```


# Get data

We read directly the data from the https source.

```{r loadFranceShapeFile}
## let's read directly the file
france_communes <- geojsonio::geojson_read(france_geojson,  what = "sp")
```

# Data check

We check that this version contains Paris intra-muros arrondissements

```{r}
sum(france_communes$code == "75115")
```


This dataset contains communes (county ?) shapefiles as well as some metadata

```{r exampleFranceShapeData}
head(france_communes@data) %>%
    kable() %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

# Create simplified versions

The dataset we obtain is quite large, as we are going to plot the whole France 
on a single map, we don’t need such resolution in the limits of communes. Thus 
we simplify the polygons with a specific algorithm.

```{r simplifyShapeFile}
france_communes_0_01 <- rmapshaper::ms_simplify(france_communes, keep = 0.01)
```

```{r}
object.size(france_communes)
object.size(france_communes_0_01)
```
 As there are no significant difference, the initial object was simplified enough.

# Create a fortified version

Let's plot France map with communes subdivisions. We need first to convert
geosjon in dataframe to be used by ggplot.

```{r FranceSimplifiedCommunes}
# Fortify the data AND keep trace of the commune code.
# communes_lim_fortified <- broom::tidy(communes_lim, region = "insee")
# broom tidy is hanging
communes_lim_fortified <- fortify(france_communes, region = "code")

```

# Plot the results

```{r}
# Now I can plot this shape easily as described before:
ggplot() +
  geom_polygon(data = communes_lim_fortified, 
               aes( x = long, y = lat, group = group),
               fill="white", 
               color="grey", size = 0.2
               ) +
  coord_map() +
  theme_tufte() +
  theme(
    axis.line=element_blank()
    , axis.text=element_blank()
    , axis.ticks=element_blank()
    , axis.title=element_blank()
    ) +
  ggtitle("France Subdivisions - Communes")
```


# Save for later use

```{r}
save(communes_lim_fortified, file = "../data/clean/communes_lim_fortified.Rda")
```


