---
title: "Get France mortality data"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
data_storage_url <- "/media/lagrange/homes/lagrange_barthelemy/_DATA_STORE/"
```

```{r loadLibraries, message=FALSE, warning=FALSE, include=FALSE}
library(data.table)
library(dplyr)
library(kableExtra)
library(ggplot2)
library(ggthemes)
```



# Introduction

The purpose of this report is to prepare a dataset containing France 
mortality data on a daily basis and per communes. 

We will combine recently published data by insee.fr and they historical data.

# Data Sources

We will use two sources of data :

- Latest official death data from french statistical bureau (https://insee.fr)
for 2020.
- Historical official death data gathered by Christian Quest from
https://data.cquest.org (this file is a 1.9 GB csv with 25 million lines). This
dataset has insee for source but gather already all past data.


```{r dataSources}
# 11Mo csv file 
# https://www.insee.fr/fr/statistiques/4487988
#insee_dcd_2020_remote_url <- "https://www.insee.fr/fr/statistiques/fichier/4487988/2020-05-15_detail.zip"
#insee_dcd_2020_remote_url <- "https://www.insee.fr/fr/statistiques/fichier/4487988/2020-05-15_detail.zip"
#insee_dcd_2020_remote_url <- "https://www.insee.fr/fr/statistiques/fichier/4487988/2020-06-19_detail.zip"
#insee_dcd_2020_remote_url <- "https://www.insee.fr/fr/statistiques/fichier/4487988/2020-12-11_detail.zip"
insee_dcd_2021_remote_url <- "https://www.insee.fr/fr/statistiques/fichier/4487988/2021-04-02_detail.zip"
```


# 2020 Deaths from Insee

## Get data

```{r getInsee2020}
insee_dcd_file <- paste0(
  data_storage_url
  , "insee.fr/mortality/raw/DC_20202021_det.csv"
  )
dcd_2020 <- data.table::fread(insee_dcd_file)
```


```{r overView2020mortality}
head(dcd_2020) %>%
    kable() %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

## Clean data

We map these data using same names as historical data (see later section)

```{r clean2020Mortality}
dcd_2020_clean <- copy(dcd_2020)
dcd_2020_clean <- dcd_2020[ , ':=' (
    sexe = SEXE
    , code_lieu_deces = COMDEC
    , date_deces = as.Date(paste0(ADEC,"-",MDEC,"-",JDEC))
    , date_naissance = as.Date(paste0(ANAIS,"-",MNAIS,"-",JNAIS))
  )]
dcd_2020_clean[ , c(
  #"DEPDOM",
  "ADEC", "MDEC", "JDEC", "DEPDEC", "ANAIS", "MNAIS", "JNAIS", "SEXE",
  "COMDEC", "COMDOM"
  , "LIEUDEC2" # could be interesting in a later stage
  ) := NULL ]
dcd_2020_clean[ sexe == "F", sexe := "2" ]
dcd_2020_clean[ sexe == "M", sexe := "1" ]
dcd_2020_clean[ , sexe := as.numeric(sexe)]
  
```

## Check & explore data

Compute daily death.

```{r dailyDeath2020}
daily_death_2020 <- dcd_2020_clean[ , .N, by = date_deces ]
```


```{r explore2020Mortality}
g <- ggplot(data = daily_death_2020)
g <- g + geom_line(mapping = aes(x = date_deces, y = N))
g <- g + ggtitle(label = "Daily death in France - 2020 data")
g <- g + xlab(label = "Date") + ylab(label = "Daily death")
g
```

These recent data contains deaths even from 2018, we will have to take this
into account while binding with archived data.

 # Historical Deaths from Insee

```{r getHistoricalMorality}
#dcd_remote_url <- "http://data.cquest.org/insee_deces/insee_deces.csv.gz"
#download.file(dcd_remote_url,destfile="./data/insee_deces.csv.gz")
dcd_url <- paste0(
  data_storage_url
  , "insee.fr/mortality/raw/insee_deces.csv"
  )
dcd_all <- data.table::fread(file = dcd_url)
```

```{r overviewMortalityHist}
head(dcd_all) %>%
    kable() %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```


We remove unused variable for the data to be merged with previous dataset and
convert dates.

```{r cleanMoralityHist}
dcd_all_clean <- dcd_all[ , .(sexe, code_lieu_deces, date_deces, date_naissance)]
dcd_all_clean[ , ':=' (
  date_deces = as.Date(date_deces)
  , date_naissance = as.Date(date_naissance)
)]
```

# Merge as Final dataset


## Exploring data to find the right cut-off date

As we have seen above, the 2020 updated death contains data from 2018, we will plot these against the historical data for comparison to decide the cut-off date.


```{r dailyComputationBoth}
daily_death_2020 <- dcd_2020_clean[ , .N, by = date_deces ]
daily_death_hist <- dcd_all_clean[ date_deces > "2017-01-01", .N, by = date_deces ]
```


```{r CompareRecentMortality}
g <- ggplot()
g <- g + geom_line(
  data = daily_death_2020
  , mapping = aes(x = date_deces, y = N)
  , col = "blue"
  , alpha = 0.5
  )
g <- g + geom_line(
  data = daily_death_hist
  , mapping = aes(x = date_deces, y = N)
  , col = "red"
  , alpha = 0.5
  )
g <- g + ggtitle(
  label = "Daily death in France since 2018"
  , subtitle = "Comparison of historical data vs. 2020 update (blue)"
    )
g <- g + xlab(label = "Date") + ylab(label = "Daily death")
g
```

The two datasets seems to match thus we will usz 2020-01-01 as cut-off date.


## Merge data

Now we merge with recent data (removing recent months)

```{r mergeMortalityDatasets}
dcd_all_clean <- rbindlist(
  list(
    dcd_all_clean[ date_deces < "2020-01-01"]
    , dcd_2020_clean[ date_deces >= "2020-01-01"]
    )
  )
```



# Explore merge data


```{r computeDailyDeathAll}
daily_death <- dcd_all_clean[ , .N, by = date_deces ]
```


```{r exploreSinceAll}
g <- ggplot(data = daily_death)
g <- g + geom_line(mapping = aes(x = date_deces, y = N))
g <- g + ggtitle(label = "Daily death in France")
g <- g + xlab(label = "Date") + ylab(label = "Daily death")
g
```

```{r explore89-90}
g <- ggplot(data = daily_death[ 
  date_deces > as.Date("1989-01-01") &
    date_deces < as.Date("1990-02-01") 
  ])
g <- g + geom_line(mapping = aes(x = date_deces, y = N))
g <- g + ggtitle(
  label = "Daily death in France - 1989 / 1990"
  ,subtitle = "Observe peak in 1989/12 and drop in 1990/01"
    )
g
```


# Save data

```{r saveData}
save(dcd_all_clean
     , file = paste0(
       data_storage_url
       , "insee.fr/mortality/clean/dcd_all_clean.Rda")
     )
```

