---
title: "Mortality in France, major events"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r loadLibraries, message=FALSE, warning=FALSE, include=FALSE}
library(data.table)
library(dplyr)
library(kableExtra)
```



# Introduction

The purpose of this report is to prepare a dataset containing France 
communes information, especially their insee code and zip code as well as
their department and regions.


# Data sources

In order to creat this dataset we will combine several sources. First from
insee.fr we will get official communes, departements and region lists with 
their insee codes and names ([insee.fr](https://www.insee.fr/fr/information/3720946) ).

We will combine these data and add the official zip codes from La Poste (French
Mail).

```{r dataSources}
# https://www.insee.fr/fr/statistiques/fichier/3720946/communes-01012019-csv.zip
# https://www.insee.fr/fr/statistiques/fichier/3720946/departement2019-csv.zip
# https://www.insee.fr/fr/statistiques/fichier/3720946/region2019-csv.zip
# https://datanova.legroupe.laposte.fr/explore/dataset/laposte_hexasmal/download/?format=csv&timezone=Europe/Berlin&lang=fr&use_labels_for_header=true&csv_separator=%3B
```


# Get data

```{r loadInseeMapping}
communes_insee <- fread(file = "../data/communes-01012019.csv")
departements_insee <- fread(file = "../data/departement2019.csv")
regions_insee <- fread(file = "../data/region2019.csv")
code_postaux <- fread(file = "../data/code_postaux.csv")
```

# Clean data 

## Insee data

We clean data from communes dataset. Basically, we removes communes which have
a parent of same code and no departement mapping. These are kind of sub communes.

```{r cleanCommunesInsee}
communes_insee <- communes_insee[ ! (dep == "" & is.na(reg))]
communes_insee[ , ':=' (
  commune_type = typecom
  , commune_insee_code = com
  , commune_name = ncc
  , commune_parent_insee_code = comparent
)]
communes_insee[ , 
                c(
                  "nccenr", "libelle", "can", 
                  "tncc", "arr", "typecom", 
                  "com", "ncc", "comparent"
                  ) := NULL 
                ]
```


```{r cleanDepartmentInsee}
departements_insee[ , ':=' (
  departement_name = ncc
  , department_cheflieu_insee_code = cheflieu
)]
departements_insee[ , 
                c(
                  "nccenr", "libelle", "tncc", "ncc", "cheflieu"
                  ) := NULL ]

```

```{r cleanRegionInsee}
regions_insee[ , ':=' (
  region_name = ncc
  , region_cheflieu_insee_code = cheflieu
)]
regions_insee[ , 
                c(
                  "nccenr", "libelle", "tncc", "cheflieu", "ncc"
                  ) := NULL ]

```

## Zip codes

```{r cleanCodePostaux}
code_postaux <- code_postaux[ , .(code_commune_insee, code_postal)]
```

# Merge datasets

We merge these 4 datasets on a single. 

```{r mapCommunesToDepartmentRegion}
communes_mapping <- merge(x = communes_insee, y = departements_insee, by.x = c("dep","reg"), by.y = c("dep","reg"))
communes_mapping <- merge(x = communes_mapping, y = regions_insee, by.x = "reg", by.y = "reg")
communes_mapping <- merge(x = communes_mapping, y = code_postaux, by.x = "commune_insee_code", by.y = "code_commune_insee")
communes_mapping[ , ':='(
  region_insee_code = reg 
  , departement_insee_code = dep
)]
communes_mapping[ , c("reg", "dep") := NULL]
communes_mapping <- communes_mapping[ , .(
  commune_insee_code, code_postal, commune_name, commune_parent_insee_code, commune_type,
  departement_insee_code, departement_name, department_cheflieu_insee_code,
  region_insee_code, region_name, region_cheflieu_insee_code
)]
```




# Explore data

```{r exampleFranceShapeData}
head(communes_mapping) %>%
    kable() %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

# Save data

```{r}
save(communes_mapping, file = "../data/clean/communes_mapping.Rda")
```

