---
title: "Analysing COVID-19 (2019-nCoV) outbreak in France"
draft: true
description: "A further exploration of COVID-19 incidence data using R tools and packages."
categories:
  - R
  - "COVID-19"
author:
  - name: Barthélémy Longueville for datapleth.io
creative_commons: CC BY-SA
twitter:
  creator: "@datapleth"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
    toc_depth: 3
    self_contained: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Analysing COVID-19 (2019-nCoV) outbreak in France}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


  
```{r loadLibs, include=FALSE}
library(tidyverse)
library(data.table)
library(kableExtra)
library(DT)
library(RColorBrewer)
library(ggplot2)
library(gghighlight)
library(plotly)
library(patchwork)
library(shadowtext)
library(viridis)
library(gt)
library(deSolve)
library(EpiEstim)
library(incidence)
library(distcrete)
library(epitrix)
library(projections)
library(earlyR)
library(distill)
library(Rovid19)

knitr::opts_chunk$set(fig.width=9, fig.height=9, echo=FALSE, warning = FALSE, message = FALSE)
# Remove scientific notation
options(scipen=999)
```




# Introduction & Objectives

This documents provides an analysis of Covid-19 outbreak in France based on 
offical data, and community data. We also include some comparison with other
countries such as Italy.

We will use different types and granularity of data such as :

- number of covid-19 case at country level
- official hospital daily data (back home, currently hospitalized, currently in ICU & deaths), at country, region, department levels
- officiel emergency daily data (ER visits, ER visits for covid, hospitalization for covid-19, SOS médecins act, SOS médecins covid-19 acts) at country, regional and department levels.
- official tests done at country, region and department levels.

Another report of this project is focusing on mortality (see references).

Visualisation and some elements of code are inspired by the best reference we
found during this period :

- John Burn-Murdoch ( @jburnmurdoch ) who is a senior dataviz journalist at ft.com (see [his work here](https://www.ft.com/coronavirus-latest))
- Tim Churches with scientist at South Western Sydney Clinical School, UNSW 
Medicine & Ingham Institute of Applied Medical Research, Liverpool, Sydney, 
Australia. ( [link](https://swscs.med.unsw.edu.au) and his 
work [here](https://timchurches.github.io/blog/))

This report is published under creative common CC BY-SA, and code should be
GPL 3.0



# Covid-19 Data per iris zone

## Covid Data

```{r}
load(file = "./latest/latest_rovid.Rda")
```


## SI-DEP geospatial data

Load Iris sub department geo file

```{r loadIris}
load(file = "/media/lagrange/homes/lagrange_barthelemy/_DATA_STORE/geo/france/clean/iris_sp.Rda")
```


## Overview

```{r}
my_date <- max(france_official_sidep_sg_iris$date, na.rm = TRUE)
france_official_sidep_sg_iris_latest <- france_official_sidep_sg_iris[ date == my_date, ]
```

The latest available data ends on `r my_date`.

## Merge data

```{r}
iris_sp_latest <- merge(
  x = iris_sp, 
  y = france_official_sidep_sg_iris_latest,
  by.x = "CODE_IRIS",
  by.y = "iris2019",
  all.x = TRUE)
```

# Let's map !

## Choropleth for 92

```{r}
my_data <- iris_sp_latest[ iris_sp_latest$DEP %in% c("59","69","43"),]
g <- ggplot(data = my_data)
g <- g + geom_sf(aes(fill = ti), lwd = 0.1) +
    scale_fill_viridis_c(option = "plasma", limits = c(0, 1000))
g 

```



