---
title: "Mortality in France, major events"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r loadLibraries, message=FALSE, warning=FALSE, include=FALSE}
library(data.table)
library(dplyr)
library(geojsonio) # require protonlite install deb : libprotobuf-dev
library(geojsonsf)
library(mapproj)
library(kableExtra)
# Need a recent version of V8
# https://github.com/jeroen/V8
# Ubuntu Xenial (16.04) and Bionic (18.04) only
# sudo add-apt-repository ppa:cran/v8
# sudo apt-get update
# sudo apt-get install libnode-dev
#devtools::install_github("jeroen/V8")
library(rmapshaper)
library(broom)
library(ggplot2)
library(ggthemes)
```



# Introduction

The purpose of this article is to analyse mortality in France over time and
over french territory.

This work was inspired by several articles (see references section) and tweets.

The main one being the tweet of @coulmont comparing the current impact of 
covid-19 on french mortality and the 2003 heatwave which killed 14k people.

<center>
<blockquote class="twitter-tweet"><p lang="fr" dir="ltr">Le &quot;pic&quot; était moins élevé en 2020 qu&#39;en 2003, mais l&#39;aire sous la courbe (c&#39;est à dire les décès en plus de la moyenne) sont déjà deux fois plus nombreux... <a href="https://t.co/r0gImxUkTj">pic.twitter.com/r0gImxUkTj</a></p>&mdash; coulmont (@coulmont) <a href="https://twitter.com/coulmont/status/1256134197217046528?ref_src=twsrc%5Etfw">May 1, 2020</a></blockquote> <script async src="https://platform.twitter.com/widgets.js" charset="utf-8"></script>
</center>

The second one is from @jburnmurdoch and compares trend accross countries.

<blockquote class="twitter-tweet"><p lang="en" dir="ltr">NEW: a lot of data on reported Covid deaths is highly suspect, so we’ve been looking into excess mortality — how many more people than usual have been dying around the world in recent weeks?<br><br>Story by me, <a href="https://twitter.com/ChrisGiles_?ref_src=twsrc%5Etfw">@ChrisGiles_</a> &amp; <a href="https://twitter.com/valentinaromei?ref_src=twsrc%5Etfw">@valentinaromei</a> (free to read): <a href="https://t.co/EiE5Q3OSmR">https://t.co/EiE5Q3OSmR</a> <a href="https://t.co/AiTdBnBma9">pic.twitter.com/AiTdBnBma9</a></p>&mdash; John Burn-Murdoch (@jburnmurdoch) <a href="https://twitter.com/jburnmurdoch/status/1254461123753054209?ref_src=twsrc%5Etfw">April 26, 2020</a></blockquote> <script async src="https://platform.twitter.com/widgets.js" charset="utf-8"></script>

We will complement this time view by geographical view, visualizing the spatial
spread of mortality overtime.

# Get & clean data

## Data sources {.tabset .tabset-face .tabset-pills}

We will use three sources of data


- Latest official death data from french statistical bureau (https://insee.fr)
for 2020.
- Historical official death data gathered by Christian Quest from
https://data.cquest.org (this file is a 1.9 GB csv with 25 million lines)
- as the data above are quite granular (individual death) and associated to 
insee code of the place when death happened, we need the mappig between insee
code and postal codes.
- finally, to map data, we need France spacial shapefile at city level.

```{r dataSources}
# 64Mo csv file
insee_dcd_all_url <- "http://data.cquest.org/insee_deces/insee_deces.csv.gz"
# 1.9Mo csv file 
insee_dcd_2020_remote_url <- "https://www.insee.fr/fr/statistiques/fichier/4470857/2020-04-30_detail.zip"
# France shapefile from data.gouv.fr
france_sp_url <- "https://data.datapleth.io/ext/france/spatial/communes-simple/communes-20190101.json"
## Important to use this version to get communes inside Paris & Marseilles.
france_geojson <- "https://github.com/gregoiredavid/france-geojson/raw/v2.0.2/communes.geojson"
# France insee code, zip code mapping
idf_insee_zip_url <- "https://www.data.gouv.fr/en/datasets/r/6d3428b2-3893-45a1-b404-2522a4e77d41"
```


## Get data

### France shapefile

```{r loadFranceShapeFile}
## let's read directly the file
france_communes <- geojsonio::geojson_read(france_geojson,  what = "sp")
```

Check that this version contains Paris intra-muros arrondissements

```{r}
sum(france_communes$code == "75115")
```


This dataset contains communes (county ?) shapefiles as well as some metadata

```{r exampleFranceShapeData}
head(france_communes@data) %>%
    kable() %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```



This data contains as well french caribeans and other oversea territories which
can be filtered out as a preliminary approach.

```{r removeFranceShapeDOMTOM}
communes_lim <- france_communes[ ! substr(france_communes@data$code,1,2)  %in% c(
    "97","98","99"
    ), ]
```


The dataset we obtain is quite large, as we are going to plot the whole France 
on a single map, we don’t need such resolution in the limits of communes. Thus 
we simplify the polygons with a specific algorithm.

```{r simplifyShapeFile}
communes_lim <- rmapshaper::ms_simplify(communes_lim, keep = 0.02)
```


Let's plot France map with communes subdivisions. We need first to convert
geosjon in dataframe to be used by ggplot.

```{r FranceSimplifiedCommunes}
# Fortify the data AND keep trace of the commune code.
# communes_lim_fortified <- broom::tidy(communes_lim, region = "insee")
# broom tidy is hanging
communes_lim_fortified <- fortify(communes_lim, region = "code")

```

```{r}
# Now I can plot this shape easily as described before:
ggplot() +
  geom_polygon(data = communes_lim_fortified, 
               aes( x = long, y = lat, group = group),
               fill="white", 
               color="grey", size = 0.2
               ) +
  coord_map() +
  theme_tufte() +
  theme(
    axis.line=element_blank()
    , axis.text=element_blank()
    , axis.ticks=element_blank()
    , axis.title=element_blank()
    ) +
  ggtitle("France Subdivisions - Communes")
```

```{r cleanObjects}
rm(france_communes)
rm(communes_lim)
```


### Death data


#### 2020 Deaths from Insee

```{r getInsee2020}
insee_dcd_file <- "../data/DC_jan2018-avr2020_det.csv"
dcd_2020 <- fread(insee_dcd_file)
```

```{r}
head(dcd_2020) %>%
    kable() %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

We map these data using same names as historical data (see later section)

```{r}
dcd_2020_clean <- dcd_2020[ , ':=' (
    sexe = SEXE
    , code_lieu_deces = COMDEC
    , date_deces = as.Date(paste0(ADEC,"-",MDEC,"-",JDEC))
    , date_naissance = as.Date(paste0(ANAIS,"-",MNAIS,"-",JNAIS))
  )]
dcd_2020_clean[ , c(
  "DEPDOM", "ADEC", "MDEC", "JDEC", "DEPDEC", "ANAIS", "MNAIS", "JNAIS", "SEXE",
  "COMDEC"
  , "LIEUDEC2" # could be interesting in a later stage
  ) := NULL ]
dcd_2020_clean[ sexe == "F", sexe := "2" ]
dcd_2020_clean[ sexe == "M", sexe := "1" ]
dcd_2020_clean[ , sexe := as.numeric(sexe)]
  
```


Compute daily death.

```{r}
daily_death_2020 <- dcd_2020_clean[ , .N, by = date_deces ]
```


```{r}
g <- ggplot(data = daily_death_2020[ date_deces >= as.Date("2020-01-01")])
g <- g + geom_line(mapping = aes(x = date_deces, y = N))
g
```




#### 2019 Deaths from Insee

```{r getInsee2019}
insee_dcd_file <- "../data/deces-2019.csv"
dcd_2019 <- fread(insee_dcd_file)
```

```{r}
dcd_2019_clean <- dcd_2019[ , ':=' (
    code_lieu_deces = lieudeces
    , date_deces = anytime::anydate(datedeces)
    , date_naissance = anytime::anydate(datenaiss)
  )]
dcd_2019_clean[ , c(
  "nomprenom", "lieudeces", "lieunaiss", "datedeces", "datenaiss", "commnaiss", "paysnaiss", "actedeces"
  ) := NULL ]
```


Compute daily death.

```{r}
daily_death_2019 <- dcd_2019_clean[ , .N, by = date_deces ]
```


```{r}
g <- ggplot(data = daily_death_2019[ date_deces > as.Date("2019-01-01")])
g <- g + geom_line(mapping = aes(x = date_deces, y = N))
g
```


#### 2019 & 2020



```{r}
dcd_all_clean2 <- rbindlist(list(
  dcd_2019_clean[date_deces < "2019-12-01"]
  , dcd_2020_clean[date_deces >= "2019-12-01"]
  )
  )
daily_death_2019_2020 <- dcd_all_clean2[ , .N, by = date_deces ]

g <- ggplot(data = daily_death_2019_2020[ date_deces > as.Date("2019-01-01")])
g <- g + geom_line(mapping = aes(x = date_deces, y = N))
g
```





#### Historical Deaths from Insee

```{r}
#dcd_remote_url <- "http://data.cquest.org/insee_deces/insee_deces.csv.gz"
#download.file(dcd_remote_url,destfile="./data/insee_deces.csv.gz")
dcd_url <- "../data/insee_deces.csv"
dcd_all <- data.table::fread(file = dcd_url)
```

```{r}
head(dcd_all) %>%
    kable() %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```


We remove unused variable for the data to be merged with previous dataset and
convert dates.

```{r}
dcd_all_clean <- dcd_all[ , .(sexe, code_lieu_deces, date_deces, date_naissance)]
dcd_all_clean[ , ':=' (
  date_deces = as.Date(date_deces)
  , date_naissance = as.Date(date_naissance)
)]
```

#### Final dataset

Now we merge with recent data (removing recent months)

```{r}
dcd_all_clean <- rbindlist(
  list(
    dcd_all_clean[ date_deces < "2019-12-01"]
    , dcd_2020_clean[ date_deces >= "2019-12-01"]
    )
  )
```

We add calendar additionnal information.

```{r}
dcd_all_clean[, ':=' (
  year_death = year(date_deces)
  , month_death = month(date_deces)
  , week_death = isoweek(date_deces)
  , dayofyear_death = as.numeric(strftime(date_deces, format = "%j"))
  , common_date_death = as.Date(paste0("2000-",format(date_deces, "%j")), "%Y-%j")
)
]
```


#### Explore historical data


```{r}
daily_death <- dcd_all_clean[ , .N, by = date_deces ]
```


```{r}
g <- ggplot(data = daily_death[ 
  #date_deces > as.Date("2000-01-01")
  ])
g <- g + geom_line(mapping = aes(x = date_deces, y = N))
g
```

```{r}
g <- ggplot(data = daily_death[ 
  date_deces > as.Date("1989-01-01") &
    date_deces < as.Date("1990-02-01") 
  ])
g <- g + geom_line(mapping = aes(x = date_deces, y = N))
g
```




### Insee code mapping 

#### France complete

We use the official mappings from [insee.fr](https://www.insee.fr/fr/information/3720946).

```{r loadInseeMapping}
# https://www.insee.fr/fr/statistiques/fichier/3720946/communes-01012019-csv.zip
# https://www.insee.fr/fr/statistiques/fichier/3720946/departement2019-csv.zip
# https://www.insee.fr/fr/statistiques/fichier/3720946/region2019-csv.zip
# https://datanova.legroupe.laposte.fr/explore/dataset/laposte_hexasmal/download/?format=csv&timezone=Europe/Berlin&lang=fr&use_labels_for_header=true&csv_separator=%3B
communes_insee <- fread(file = "../data/communes-01012019.csv")
departements_insee <- fread(file = "../data/departement2019.csv")
regions_insee <- fread(file = "../data/region2019.csv")
code_postaux <- fread(file = "../data/code_postaux.csv")
```

We clean data from communes dataset. Basically, we removes communes which have
a parent of same code and no departement mapping. These are kind of sub communes.


```{r cleanCommunesInsee}
communes_insee <- communes_insee[ ! (dep == "" & is.na(reg))]
communes_insee[ , ':=' (
  commune_type = typecom
  , commune_insee_code = com
  , commune_name = ncc
  , commune_parent_insee_code = comparent
)]
communes_insee[ , 
                c(
                  "nccenr", "libelle", "can", 
                  "tncc", "arr", "typecom", 
                  "com", "ncc", "comparent"
                  ) := NULL ]

```


```{r cleanDepartmentInsee}
departements_insee[ , ':=' (
  departement_name = ncc
  , department_cheflieu_insee_code = cheflieu
)]
departements_insee[ , 
                c(
                  "nccenr", "libelle", "tncc", "ncc", "cheflieu"
                  ) := NULL ]

```

```{r cleanRegionInsee}
regions_insee[ , ':=' (
  region_name = ncc
  , region_cheflieu_insee_code = cheflieu
)]
regions_insee[ , 
                c(
                  "nccenr", "libelle", "tncc", "cheflieu", "ncc"
                  ) := NULL ]

```

```{r cleanCodePostaux}
code_postaux <- code_postaux[ , .(code_commune_insee, code_postal)]
```


We merge these 4 datasets on a single. 

```{r mapCommunesToDepartmentRegion}
communes_mapping <- merge(x = communes_insee, y = departements_insee, by.x = c("dep","reg"), by.y = c("dep","reg"))
communes_mapping <- merge(x = communes_mapping, y = regions_insee, by.x = "reg", by.y = "reg")
communes_mapping <- merge(x = communes_mapping, y = code_postaux, by.x = "commune_insee_code", by.y = "code_commune_insee")
communes_mapping[ , ':='(
  region_insee_code = reg 
  , departement_insee_code = dep
)]
communes_mapping[ , c("reg", "dep") := NULL]
communes_mapping <- communes_mapping[ , .(
  commune_insee_code, code_postal, commune_name, commune_parent_insee_code, commune_type,
  departement_insee_code, departement_name, department_cheflieu_insee_code,
  region_insee_code, region_name, region_cheflieu_insee_code
)]
```





#### Ile de France

We have only mapping for Ile de France, need to look for whole France.

```{r}
insee_idf_mapping <- as.data.table(read.csv(file = idf_insee_zip_url, sep = ";", stringsAsFactors = FALSE))
```

```{r}
insee_idf_mapping <- insee_idf_mapping[ , .(
  Code.INSEE, Code.Postal, 
  #Commune, 
  Département, Région,
  #Population,
  Code.Département 
)]
setnames(insee_idf_mapping, c(
  "code_insee"
  , "code_postal"
  , "departement_name"
  , "region_name"
  , "code_departement"
))
```

Update variable types

```{r}
insee_idf_mapping[ , ':=' (
  code_insee = as.character(code_insee)
)]
```



```{r}
head(insee_idf_mapping) %>%
    kable() %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```


## Create summarized data

### Total of death per day and insee_code (county/city)

#### Agregate per day

```{r}
dcd_by_code_daily <- copy(dcd_all_clean)
dcd_by_code_daily <- dcd_by_code_daily[ , .(count = .N), by = .(code_lieu_deces,date_deces,year_death,month_death,week_death, dayofyear_death,common_date_death) ]
# convert to date
#dcd_by_code_daily_idf[ , date_deces := as.Date (date_deces)]
```


#### Merge with insee mapping

```{r}
dcd_by_code_daily <- merge(
  x = dcd_by_code_daily
  , y = communes_mapping
  , by.x = "code_lieu_deces"
  , by.y = "commune_insee_code"
  , allow.cartesian = TRUE
  )
```

```{r}
head(dcd_by_code_daily) %>%
    kable() %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```



#### Visualisation

```{r}
dcd_by_code_daily_s <- dcd_by_code_daily[ year_death > 1972 ]
dcd_by_code_daily_s <- dcd_by_code_daily_s[ , .( sum_c = sum(count) ) , by = .(year_death,common_date_death)] 

```


```{r}
library(gghighlight)
g <- ggplot(data = dcd_by_code_daily_s, group = year_death)
g <- g + geom_line(mapping = aes(x = common_date_death, y = sum_c, col = as.factor(year_death)), size = 0.5)
g <- g + gghighlight(year_death %in% c(2020,2003,1989, 1994, 1997,1970))
g <- g + scale_x_date(labels = function(x) format(x, "%d-%b")) 
g <- g + ggtitle("Mortalité quotidienne par année - France*** ")
g <- g + xlab("Jour") + ylab("nb de décès")
g <- g + theme_minimal()
g
```


#### Choropleth for a specific date

```{r}
communes_results <- merge(
  x = dcd_by_code_daily[ date_deces == "2020-04-01"]
  , y = communes_lim_fortified
  , by.x = "code_lieu_deces"
  , by.y = "id"
  , all.y = TRUE
  , allow.cartesian=TRUE
)
```

```{r}
p <- ggplot() +
  geom_polygon(
    data = communes_results, 
    aes(fill = count, x = long, y = lat, group = group),
    color="grey", size = 0.02
  ) +
  scale_fill_viridis_c(option = "A", direction = -1) +
  #facet_wrap(facets = . ~ liste_shortname) +
  coord_map() +
  theme_tufte() +
  theme(
    axis.line=element_blank()
    , axis.text=element_blank()
    , axis.ticks=element_blank()
    , axis.title=element_blank()
    ) +
  ggtitle("European election - 2019 - France") +
  labs(fill = "nb_death")
p
```




### Total of death per week and insee_code (county/city)

#### Agregate per week

```{r}
dcd_by_code_weekly <- copy(dcd_all_clean)
dcd_by_code_weekly <- dcd_by_code_weekly[ , .(count = .N), by = .(code_lieu_deces,week_death,year_death,month_death) ]
```


#### Merge with insee mapping

```{r}
dcd_by_code_weekly <- merge(
  x = dcd_by_code_weekly
  , y = communes_mapping
  , by.x = "code_lieu_deces"
  , by.y = "commune_insee_code"
  , allow.cartesian = TRUE
  )
```

```{r}
head(dcd_by_code_weekly) %>%
    kable() %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```



#### Visualisation

```{r}
dcd_by_code_weekly_s <- dcd_by_code_weekly[ year_death > 1972 ]
dcd_by_code_weekly_s <- dcd_by_code_weekly_s[ , .( sum_c = sum(count) ) , by = .(year_death,week_death)] 

```


```{r}
library(gghighlight)
g <- ggplot(data = dcd_by_code_weekly_s, group = year_death)
g <- g + geom_line(mapping = aes(x = week_death, y = sum_c, col = as.factor(year_death)), size = 0.5)
g <- g + gghighlight(year_death %in% c(2020,2003,1989, 1994, 1997,1970))
#g <- g + scale_x_date(labels = function(x) format(x, "%d-%b")) 
g <- g + ggtitle("Mortalité quotidienne par année - France*** ")
g <- g + xlab("Week") + ylab("nb de décès")
g <- g + theme_minimal()
g
```


#### Choropleth for a specific week

```{r}
communes_results_w <- merge(
  x = dcd_by_code_weekly[ week_death == 12]
  , y = communes_lim_fortified
  , by.x = "code_lieu_deces"
  , by.y = "id"
  , all.y = TRUE
  , allow.cartesian=TRUE
)
```

```{r}
p <- ggplot() +
  geom_polygon(
    data = communes_results_w[ 
      departement_insee_code %in% c(
      #"92","91","93", "94","95","78", "75"
      "69"
      # , "69", "13","33","19"
      ) | TRUE
      ], 
    aes(fill = count, x = long, y = lat, group = group),
    color="grey", size = 0.02
  ) +
  scale_fill_viridis_c(option = "A", direction = -1) +
  #facet_wrap(facets = . ~ liste_shortname) +
  coord_map() +
  theme_tufte() +
  theme(
    axis.line=element_blank()
    , axis.text=element_blank()
    , axis.ticks=element_blank()
    , axis.title=element_blank()
    ) +
  ggtitle("Weekly mortality - 2020 - Week 12 - France") +
  labs(fill = "nb_death")
p
```




# References


https://freakonometrics.hypotheses.org/60845