---
title: "Mortality in France, major events"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
Sys.setlocale(category="LC_ALL", locale="C")
## configure path for data storage
data_storage_url <- "/media/lagrange/homes/lagrange_barthelemy/_DATA_STORE/"
```

```{r loadLibraries, message=FALSE, warning=FALSE, include=FALSE}
library(data.table)
library(dplyr)
library(geojsonio) # require protonlite install deb : libprotobuf-dev
library(geojsonsf)
library(mapproj)
library(kableExtra)
# Need a recent version of V8
# https://github.com/jeroen/V8
library(rmapshaper)
library(broom)
library(ggplot2)
library(ggthemes)
library(gghighlight)
```


# Introduction & Objective {.tabset .tabset-face .tabset-pills}

## Inspiration

The purpose of this article is to analyse mortality in France over time and
over french territory at different scales.

This work was inspired by several articles (see references section) and tweets.

The main one being the tweet of @coulmont comparing the current impact of 
covid-19 on french mortality and the 2003 heatwave which killed 14k people.

<center>
<blockquote class="twitter-tweet"><p lang="fr" dir="ltr">Le &quot;pic&quot; était moins élevé en 2020 qu&#39;en 2003, mais l&#39;aire sous la courbe (c&#39;est à dire les décès en plus de la moyenne) sont déjà deux fois plus nombreux... <a href="https://t.co/r0gImxUkTj">pic.twitter.com/r0gImxUkTj</a></p>&mdash; coulmont (@coulmont) <a href="https://twitter.com/coulmont/status/1256134197217046528?ref_src=twsrc%5Etfw">May 1, 2020</a></blockquote> <script async src="https://platform.twitter.com/widgets.js" charset="utf-8"></script>
</center>

The second one is from @jburnmurdoch and compares trend accross countries.

<center>
<blockquote class="twitter-tweet"><p lang="en" dir="ltr">NEW: a lot of data on reported Covid deaths is highly suspect, so we’ve been looking into excess mortality — how many more people than usual have been dying around the world in recent weeks?<br><br>Story by me, <a href="https://twitter.com/ChrisGiles_?ref_src=twsrc%5Etfw">@ChrisGiles_</a> &amp; <a href="https://twitter.com/valentinaromei?ref_src=twsrc%5Etfw">@valentinaromei</a> (free to read): <a href="https://t.co/EiE5Q3OSmR">https://t.co/EiE5Q3OSmR</a> <a href="https://t.co/AiTdBnBma9">pic.twitter.com/AiTdBnBma9</a></p>&mdash; John Burn-Murdoch (@jburnmurdoch) <a href="https://twitter.com/jburnmurdoch/status/1254461123753054209?ref_src=twsrc%5Etfw">April 26, 2020</a></blockquote> <script async src="https://platform.twitter.com/widgets.js" charset="utf-8"></script>
</center>

We will complement this time view by geographical view, visualizing the spatial
spread of mortality overtime.

## Data sources

We are combining five resources :

- mortality daily data per communes based on latest official death data from 
french statistical bureau (https://insee.fr) for 2020 and historical official
death data gathered by Christian Quest from https://data.cquest.org from
insee.fr as well.
- French population over time per departements (from https://insee.fr), prepared
as well in a separated notebook.
- geographic spatial limits of communes fortified as a data frame to be plotted
with ggplot (from https://github.com/gregoiredavid/france-geojson)
- geographic spatial limits of departement fortified as a data frame to be 
plotted with ggplot (from https://github.com/gregoiredavid/france-geojson)
- the list of all french communes with their insee codes and zip codes, this is
a merge from La Poste open data and insee.fr


## Load data

We prepared in separated notebook the datasets mentioned previously, thus we
just have to load R objects here.


```{r loadDataFromFolder, message=FALSE, warning=FALSE}
# shapefiles for communes and departements
load(
  file = paste0(
    data_storage_url
    , "geo/france/clean/communes_lim_fortified.Rda"
    )
  )
load(
  file = paste0(
    data_storage_url
    , "geo/france/clean/departements_lim_fortified.Rda"
    )
  )
# List of Communes and metadata such as zip codes
load(
  file = paste0(
    data_storage_url
    , "insee.fr/codes/clean/communes_mapping.Rda"
    )
)
# All individual deaths since 1970 in France
load(
  file = paste0(
    data_storage_url
    , "insee.fr/mortality/clean/dcd_all_clean.Rda"
    )
)
# France_population_per_departement and years
load(
  file = paste0(
    data_storage_url
    , "insee.fr/population/raw/estim-pop-dep-sexe-gca-1975-2020.Rda"
    )
)

```


# Mortality at country level


In this section we will analyse mortality in France at country level and
identify the main events.

## France Total deaths per day

Let's first compute the total number of death per day since 1970 and compare
trends.

Mortality data set contains all individual deaths, we just have to count these
per date of death. Then we add some useful metadata as new features to improve
data visualisation in the next sections.

```{r createDcdDaily}
dcd_daily <- copy(dcd_all_clean)
# data.table is super fast for such tasks
dcd_daily <- dcd_daily[ , .(daily_deaths = .N), by = .(date_deces) ]
# add new features
dcd_daily[, ':=' (
  year_death = year(date_deces)
  , month_death = month(date_deces)
  , week_death = isoweek(date_deces)
  , dayofyear_death = as.numeric(strftime(date_deces, format = "%j"))
  # bellow is a dummy variable for pretty prints of plots, mapping the
  # day of the year in 2000
  , common_date_death = as.Date(paste0("2000-",format(date_deces, "%j")), "%Y-%j")
)
]
```

Let's have a look at a small extract of the data.

```{r overViewDaily, echo=FALSE}
head(dcd_daily) %>%
    kable() %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

## Visualize the data

In the following plot we illustrate trends of mortality in France in number of 
deaths per day for the whole country over time since 1973. We highlighted some
specific years.

```{r mortalityComparison, echo=FALSE, message=FALSE, warning=FALSE}
# Years to be highlighted
year_selected <- c(
    1989
    #, 1994, 1997,
    ,1997 ,1999, 2003
    #2017
    , 2020
    )
# Define the number of colors you want
nb_cols <- length(unique(year_selected))
mycolors <- grDevices::colorRampPalette(RColorBrewer::brewer.pal(5, "Dark2"))(nb_cols)
  
g <- ggplot(
  data = dcd_daily[ date_deces > as.Date("1973-12-31") ]
  , group = year_death
  )
g <- g + geom_line(
  mapping = aes(
    x = common_date_death
    , y = daily_deaths
    , col = as.factor(year_death)
    )
  , size = 0.8)
g <- g + gghighlight(
  year_death %in% year_selected
  , use_direct_label = FALSE
  , unhighlighted_params = list(
      size = 0.3
      , alpha = 0.5
      )
  ) + scale_color_manual(values = mycolors)
g <- g + scale_x_date(labels = function(x) format(x, "%d-%b")) 
g <- g + ggtitle("Mortality in France")
g <- g + labs(
  x = NULL
  , y = "deaths"
  , colour = NULL
  )
g <- g + ylim(c(0,4000)) + theme_hc() #theme_tufte() #theme_minimal() 
# Add annotation
g <- g + annotate(
  "text", x = as.Date("2000-08-08"), y = 3800,
  label = "Heatwave", alpha = .3, size = 3
  )
g <- g + annotate(
  "text", x = as.Date("2000-04-01"), y = 2900,
  label = "Covid-19", alpha = .3, size = 3
  )
g <- g + annotate(
  "text", x = as.Date("2000-03-01"), y = 2600,
  label = "Flu", alpha = .3, size = 3
  )
g <- g + annotate(
  "text", x = as.Date("2000-12-15"), y = 3800,
  label = "1989 Flu ?", alpha = .3, size = 3
  )
g <- g + annotate(
  "text", x = as.Date("2000-07-20"), y = 500,
  label = "June Flu + low summer ?", alpha = .3, size = 3
  )
g
```

Mortality in France has a U shape over one year with highler level during
winter. We highlighted major events :

- 2003 summer heatwave
- 2020 covid-19
- 2017 & 1999 influenza
- 1997 summer trends look awkward with a small peak of mortality in June and a
low mortality during summer
- 1989 mortality peak during winter looks very high and long. This will requires
further checking as we didn't find articles or references regarding this event.


# Mortality at *departement* level

Now that we have a global idea of mortality trends and events at country level
we will check local situation, at *departement* level.

### Prepare the data

The deaths database contains insee code for *communes*, we need to proceed in 
two steps to get agregated data at *departement* level. We ifrst compute the
daily number of deaths per location code, at communes level, then merge
with *communes* insee code mapping with *departement* numbers and finally 
agregate data at this level.

```{r createDailyDataDep1}
dcd_by_communes_daily <- copy(dcd_all_clean)
dcd_by_communes_daily <- dcd_by_communes_daily[ , .(daily_deaths = .N), by = .(code_lieu_deces,date_deces) ]

dcd_by_communes_daily <- merge(
  x = dcd_by_communes_daily
  , y = communes_mapping
  , by.x = "code_lieu_deces"
  , by.y = "commune_insee_code"
  , allow.cartesian = TRUE
  )

dcd_by_department_daily <- dcd_by_communes_daily[ , .(daily_dep_deaths = sum(daily_deaths)), by = .(region_name,departement_insee_code,date_deces) ]
```


Then we add some usefull metadata as new feature necessary in the next plots.

```{r prepareDataVizDaily1}
dcd_by_department_daily[, ':=' (
  year_death = as.factor(year(date_deces))
  , month_death = month(date_deces)
  , week_death = isoweek(date_deces)
  , dayofyear_death = as.numeric(strftime(date_deces, format = "%j"))
  , common_date_death = as.Date(paste0("2000-",format(date_deces, "%j")), "%Y-%j")
)
]
```


### Visualization

Let's check the results with a heatmap over the last two years. In this plot
we observe the intensity of deaths (in number of deaths) per day and per 
departements. As departement have not the same size, it's not possible to 
compare line by line but we can observe the covid-19 effect in the region of
Paris and east of France. The largest departement of France (59, 06) have 
obviously the largest number of deaths.

```{r}
mypal <- c("#e7f0fa", #lighter than light blue
         "#c9e2f6", #light blue
         "#95cbee", #blue
         "#0099dc", #darker blue
         "#4ab04a", #green
         "#ffd73e", #yellow
         "#eec73a", #mustard
         "#e29421", #dark khaki (?)
         "#f05336", #orange red
         "#ce472e") #red

g <- ggplot(
    data = dcd_by_department_daily[ date_deces > as.Date("2019-01-01") ]
    , aes(
      x = date_deces
      , y = departement_insee_code
      , fill = daily_dep_deaths
      )
    ) +
    geom_tile(colour="white", linejoin = 2, 
              width=.9, height=.9) + theme_minimal()  +
    scale_fill_gradientn(
        colours = mypal
        # for daily death
        , values=c(0, 0.05, 0.1, 0.15, 0.2, 0.25, .3, .35, .4, .7, 1)
        , limits=c(0, 300)
        , na.value=rgb(246, 246, 246, max=255)
        #, labels=c("0", "1", "2k", "3k", "4k")
        , guide = guide_colourbar( barwidth= 0.6 )
        ) +
    #scale_y_discrete(limits = rev(levels(droplevels(dcd_by_department_daily$departement_insee_code)))) +
   # scale_x_continuous(expand=c(0,0), 
    #                   breaks=seq(1930, 2010, by=10)) +
    #(x=1963, xend=1963, y=0, yend=56.5, size=.7) +
    ggtitle("Evolution of daily deaths per department in France") +
    theme(
        axis.title.y=element_blank()
        , axis.text.y = element_text(size = 6)
        , axis.text.x = element_text(angle = 90)
        , panel.grid = element_blank()
        , axis.ticks.y=element_blank()
        , strip.text.y.right = element_text(angle = 0)
        )
g <- g + facet_grid(facets = region_name ~ ., scales = "free_y")
g
```

## France variation of mortality per departements

### Discussion on absolute number of deaths

As discussed in the previous section, the absolute number of death is a good
indicator to visualize the effect of covid-19 in a specific departement but
cannot be used to compare departements. We could use specific mortality rate
but let's first simply compute daily deaths as percentage of the last 5 years
average for the same date in the year.

### Compute 5 years average deaths per departements.

For each departement, we will compute the daily average on the last 10 years.
Then we will smooth this value by computing a rolling average of a week.


```{r averageDeathPerDep}
dcd_by_department_average <- dcd_by_department_daily[ 
  date_deces >= as.Date("2015-01-01") &
    date_deces < as.Date("2020-01-01")
    ]
dcd_by_department_average <- dcd_by_department_average[ , .(deaths_av  = mean(daily_dep_deaths)), by = .(region_name,departement_insee_code,common_date_death) ]
dcd_by_department_average <- dcd_by_department_average[order(region_name,departement_insee_code,common_date_death)]
```

```{r}
my_smooth <- function (y){
  # build a 2d data with inputs
  data <- data.table( x = 1:length(y), y = y)
  # fit a loess with high span
  # WARNING, x must be ordered
  z <- with(data, loess(y ~ x, span = 0.4))
  # Return fitted values
  z$fitted
}


dcd_by_department_average[ , ':=' (
  deaths_av_smooth = my_smooth(deaths_av)
  )
  , by =  departement_insee_code 
  ]
```





### Visualisation 

#### Trends of mortality over the last 10 years


```{r mortalityPerDep}
dep_filter <- c(
  #"75","92", "94", "93"
  "04","06"
  #,"19", "23", "87"
  ) 
my_data_history <- dcd_by_department_average[ 
  departement_insee_code %in% dep_filter
  | TRUE
  ]


## Start the plot
g <- ggplot(data = my_data_history)
# Add mean daily values
#g <- g + geom_line(aes(x= common_date_death, y = deaths_av), col = "grey", alpha = 0.5)
# Add daily values from the last 5 years
g <- g + geom_line(
  data = dcd_by_department_daily[
    date_deces >= as.Date("2015-01-01") &
    date_deces < as.Date("2020-01-01") &
    (
      departement_insee_code %in% dep_filter
      | TRUE
    )
  ]
  , aes(x= common_date_death, y = daily_dep_deaths), col = "grey", alpha = 0.5
  )
# Add smooth of average last 5 years values
g <- g + geom_line(aes(x= common_date_death, y = deaths_av_smooth), col = "darkred", alpha = 0.8)

g <- g + facet_wrap(facets = . ~ departement_insee_code, ncol = 10, scales = "free_y")
g <- g + theme_minimal() +
    theme(
      #legend.position = "none"
      strip.text.x = element_text(
        size=5
        , face = "bold"
        , vjust = -0.0,
        margin = margin(-0.00, 0, 0, 0, "cm")
      )
      , strip.background = element_rect(
        color ="white"
        , fill ="white"
        , size = 0.5
        #, linetype="solid"
      )
      , axis.text.x = element_text(
        angle = 90
        , size = 5
      )
      , axis.text.y = element_text(size = 5)
      )
g <- g + scale_x_date(date_labels = "%b/%d")
g <- g + ggtitle(
  label = "Daily Mortality trends per departement"
  , subtitle = "2015 ~ 2019 average") +
  xlab(label = "Date") + ylab("Daily deaths")
g
```


#### Case of 2020

```{r mortalityPerDepwith2020}
# Define the number of colors you want
nb_cols <- length(unique(communes_mapping$region_name))
mycolors <- grDevices::colorRampPalette(RColorBrewer::brewer.pal(9, "Dark2"))(nb_cols)

dep_filter <- c(
  "01","02","03","04","05","06","07","08","09"
  ,as.character(10:30)
  ,as.character(31:60)
  ,as.character(61:90)
  )
my_data_2020 <- dcd_by_department_daily[
    year_death == 2020 &
    (departement_insee_code %in% dep_filter
  | TRUE )
  ]
my_data_history <- dcd_by_department_average[ 
  ( departement_insee_code %in% dep_filter
  | TRUE )
  ]


## Start the plot
g <- ggplot(data = my_data_history)
# Add mean daily values
g <- g + geom_line(
  aes(x= common_date_death, y = deaths_av)
  , col = "grey", alpha = 0.3
  )
# Add smooth of average last 5 years values
g <- g + geom_line(
  aes(x= common_date_death, y = deaths_av_smooth)
  , col = "darkgrey"
  , alpha = 0.8
  )
# Add 2020 data
g <- g + geom_line(
  data = my_data_2020
  , aes( x= common_date_death, y = daily_dep_deaths)
  , col = "darkblue", alpha = 0.2
  )

# Add 2020 data Smooth
g <- g + geom_smooth(
  data = my_data_2020
  , aes( 
    x= common_date_death
    , y = daily_dep_deaths
    #, col = region_name
    )
  , method = "loess"
  , span = 0.5
  , se = FALSE
  , col = "darkblue"
  , alpha = 0.8
  )  #+ scale_color_manual(values = mycolors)

g <- g + theme_minimal() +
    theme(
      #legend.position = "none"
      strip.text.x = element_text(
        size=5
        , face = "bold"
        , vjust = -0.0,
        margin = margin(-0.00, 0, 0, 0, "cm")
      )
      , strip.background = element_rect(
        color ="white"
        , fill ="white"
        , size = 0.5
        #, linetype="solid"
      )
      , axis.text.x = element_text(
        angle = 90
        , size = 5
      )
      , axis.text.y = element_text(size = 5)
      )
g <- g + scale_x_date(date_labels = "%b/%d") +
  expand_limits(y = 0) 
 # ylim(c(0,130)) + xlim(c(as.Date("2000-01-01"), as.Date("2000-06-15"))
  # geom_vline( xintercept = as.Date("2000-03-15"), color = "darkgrey", size= 0.6, alpha = 0.7 )
g <- g + ggtitle(
  label = "Daily Mortality in 2020 per departement"
  , subtitle = "vs. 2015 ~ 2019 average") +
  xlab(label = "Date") + ylab("Daily deaths")

h1 <- g + facet_wrap(
  facets = region_name ~ departement_insee_code
  , ncol = 10
  #, scales = "free_y"
  )
h2 <- g + facet_wrap(
  facets = region_name ~ departement_insee_code
  , ncol = 10
  , scales = "free_y"
  )
h1
h2
```



## Total of death per day and insee_code (county/city)

### Agregate per day

```{r createDailyData2}
dcd_by_code_daily <- copy(dcd_all_clean)
dcd_by_code_daily <- dcd_by_code_daily[ , .(daily_deaths = .N), by = .(code_lieu_deces,date_deces) ]
```


### Merge with insee mapping

```{r mergeDailWithInsee}
dcd_by_code_daily <- merge(
  x = dcd_by_code_daily
  , y = communes_mapping
  , by.x = "code_lieu_deces"
  , by.y = "commune_insee_code"
  , allow.cartesian = TRUE
  )
```

### Add metadata 

```{r prepareDataVizDaily2}
dcd_by_code_daily[, ':=' (
  year_death = as.factor(year(date_deces))
  , month_death = month(date_deces)
  , week_death = isoweek(date_deces)
  , dayofyear_death = as.numeric(strftime(date_deces, format = "%j"))
  , common_date_death = as.Date(paste0("2000-",format(date_deces, "%j")), "%Y-%j")
)
]
```



### Overview of the data

```{r overViewDailybyCode}
head(dcd_by_code_daily) %>%
    kable() %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

### Choropleth for a specific date

```{r}
communes_results <- merge(
  x = dcd_by_code_daily[ date_deces == "2020-04-01" ]
  , y = communes_lim_fortified
  , by.x = "code_lieu_deces"
  , by.y = "id"
  , all.y = TRUE
  , allow.cartesian=TRUE
)
```

```{r}
p <- ggplot() +
  geom_polygon(
    data = communes_results, 
    aes(fill = daily_deaths, x = long, y = lat, group = group),
    color="grey", size = 0.02
  ) +
  scale_fill_viridis_c(option = "A", direction = -1) +
  #facet_wrap(facets = . ~ liste_shortname) +
  coord_map() +
  theme_tufte() +
  theme(
    axis.line=element_blank()
    , axis.text=element_blank()
    , axis.ticks=element_blank()
    , axis.title=element_blank()
    ) +
  ggtitle("Daily mortality - 2020-04-01 - France") +
  labs(fill = "nb_death")
p
```

# References

## Inspirations 
https://freakonometrics.hypotheses.org/60845

## Data Sources


```{r dataSources}
# 64Mo csv file
insee_dcd_all_url <- "http://data.cquest.org/insee_deces/insee_deces.csv.gz"
# 1.9Mo csv file 
insee_dcd_2020_remote_url <- "https://www.insee.fr/fr/statistiques/fichier/4470857/2020-04-30_detail.zip"
# France shapefile from data.gouv.fr
france_sp_url <- "https://data.datapleth.io/ext/france/spatial/communes-simple/communes-20190101.json"
## Important to use this version to get communes inside Paris & Marseilles.
france_geojson <- "https://github.com/gregoiredavid/france-geojson/raw/v2.0.2/communes.geojson"
# France insee code, zip code mapping
idf_insee_zip_url <- "https://www.data.gouv.fr/en/datasets/r/6d3428b2-3893-45a1-b404-2522a4e77d41"
```
